MATD02 - Exploration of Finite Geometry

### 2024-11-17 paper/main.tex
Finish proof of theroem 1, try to wrap up future developments section

% Format bibliography links correctly
\usepackage[hidelinks]{hyperref}
\usepackage[T1]{fontenc}

@@

    In fact, it does generalize, we call the first concept duality,
    in which theorems about lines on a projective plane also have duals regarding points (and vice-versa).
    The second gives rise to a concept of order\cite[p. 24]{beutelspacher_projective_1998}.

    \begin{theorem}
        Given a projective space, there exists a positive integer $q$ such that any line in $L$ contains exactly $q+1$ elements.
        We call q the order of this space.
    \end{theorem}

    \begin{proof}
        Given any two lines $L_1, L_2$, we'll argue there exists a bijection $\phi: L_1 \rightarrow L_2$,
        and thus all lines have the same number of elements.
        If $L_1 = L_2$ we are done, so we consider distinct lines.

        \underline{Case 1}: $L_1$ intersects $L_2$

        Then $\{a, b\} \subseteq L_1$ and $\{a, c\} \subseteq L_2$ for some points $a,b,c$,
        since by A3 both lines contain at least 3 distinct points and by supposition we have at least one shared point.

        As well, by A3 line $bc$ must contain another point, name it $p$, moreover $p \notin L_1, L_2$,
        otherwise $L_1 = L_2$ by A1, the uniqueness of lines through points.

        Then given any other point $x \in L_1$, consider the line $xp$.
        We'll show this must intersect $L_2$ using A2 and call the intersection point $\phi(x)$.

        \begin{figure}[h]
            \centering
            \begin{tikzpicture}
                % Draw the lines
                \draw[thick] (-2,0) -- (3,1) node[right] {$L_1$};
                \draw[thick] (-2,0) -- (3,-1) node[right] {$L_2$};
                % Draw the intersection point
                \fill (-2,0) circle (2pt) node[below] {$a$};
                % Label points on L1
                \fill (2,0.8) circle (2pt) node[above] {$b$};
                \fill (0,0.4) circle (2pt) node[above] {$x$};
                % Label points on L2
                \fill (2,-0.8) circle (2pt) node[below] {$\phi(x)$};
                \fill (0,-0.4) circle (2pt) node[below] {$c$};
                % Draw intersecting lines
                \draw[thick] (2,-0.8) -- (0,0.4)
                \draw[thick] (2,0.8) -- (0,-0.4)
                \fill (0.7,0) circle (2pt) node[below] {$p$};
            \end{tikzpicture}
%            \caption{Illustration of our points and lines}
            \label{fig:intersecting_lines}
        \end{figure}

        Given points $a,b,c,d$, if $ab$ intersects $cd$, then $ac$ intersects $bd$

        \underline{Case 2}: $L_1$ does not intersect $L_2$
    \end{proof}

@@

    The Veblen-Young theorem shows finite projective spaces of dimension $n \geq 3$ can always be described using such a field.
    However, finite projective planes are much harder to classify, since they can be constructed in more ways.
    Specifically, non-Desarguesian planes.
    The difficulty in classifying finite projective planes gives rise to an open question:
    \textit{Is the order of a finite projective plane always a prime power?}
    \todo{Mention examples? State Desargues' theorem?}

    The number of non-isomorphic finite projective planes is only known up to order 10: $1, 1, 1, 1, 0, 1, 1, 4, 0$
    due to the computational complexity of verifying this using brute force\cite{lam_computer_1991,lam_non-existence_1989}.
    Moreover, the closest we have come to a proof of this conjecture is the Bruck-Ryser theorem, which states if $n \equiv 1,2\ (mod\ 4)$ and $n$ is not the sum of two squares,
    then no plane exists of order $n$.

    \todo{Bruck–Ryser theorem}
    \todo{Classification up to order 10, why is it so difficulty to do past this small number?}

>>>>

% Format bibliography links correctly
\usepackage[hidelinks]{hyperref}
\usepackage[T1]{fontenc}
\usepackage{mathtools}

@@

    In fact, it does generalize, we call the first concept duality,
    in which theorems about lines on a projective plane also have duals regarding points (and vice-versa).
    In finite projective spaces, the second gives rise to a concept of order\cite[p. 24]{beutelspacher_projective_1998}.

    \begin{theorem}
        Given a finite projective space, there exists a positive integer $q$ such that any line in $L$ contains exactly $q+1$ elements.
        We call q the order of this space.
    \end{theorem}

    \begin{proof}
        Given any two lines $L_1, L_2$, we'll argue there exists a bijection $\phi: L_1 \rightarrow L_2$,
        and thus all lines have the same number of elements.
        When our set of points is finite, this implies the existence of an integer order.
        If $L_1 = L_2$ we are done, so we consider distinct lines.

        \underline{Case 1}: $L_1$ intersects $L_2$.
        Then $\{a, b\} \subseteq L_1$ and $\{a, c\} \subseteq L_2$ for some points $a,b,c$,
        since by A3 both lines contain at least 3 distinct points and by supposition we have at least one shared point.
        By A3 again, line $bc$ must contain another point, name it $p$, moreover $p \notin L_1, L_2$,
        otherwise $L_1 = L_2$ by A1, the uniqueness of lines through points.

        Then given any other point $x \in L_1$, consider the line $xp$.
        We'll show this must intersect $L_2$ and call the intersection point $\phi(x)$.

        \begin{figure}[h]
            \centering
            \begin{tikzpicture}
                % Draw the lines
                \draw[thick] (-2,0) -- (3,1) node[right] {$L_1$};
                \draw[thick] (-2,0) -- (3,-1) node[right] {$L_2$};
                % Draw the intersection point
                \fill (-2,0) circle (2pt) node[below] {$a$};
                % Label points on L1
                \fill (2,0.8) circle (2pt) node[above] {$b$};
                \fill (0,0.4) circle (2pt) node[above] {$x$};
                % Label points on L2
                \fill (2,-0.8) circle (2pt) node[below] {$\phi(x)$};
                \fill (0,-0.4) circle (2pt) node[below] {$c$};
                % Draw intersecting lines
                \draw[thick] (2,-0.8) -- (0,0.4)
                \draw[thick] (2,0.8) -- (0,-0.4)
                \fill (0.7,0) circle (2pt) node[below] {$p$};
            \end{tikzpicture}
%            \caption{Illustration of our points and lines}
            \label{fig:intersecting_lines}
        \end{figure}
        By A2: If $xa$ intersects $pc$, then $xp$ intersects $ac$.
        We have $pc=bc$ and $xa = ab$ by A1, so both lines contain $b$ and thus intersect.
        So $xp$ intersects $ac$.
        We define the function $\phi(x)$ with domain $L_1$ as follows:

        \[
            \phi(x) =
            \begin{cases}
                xp \cap L_2 & \text{if } x \neq a \\
                a & \text{if } x = a
            \end{cases}
        \]

        Let $x_2\in L_1$ be distinct from $a, x$.
        Then $\phi(x_2) \neq \phi(x)$,
        since both pass through p and otherwise by A1 we would have $x = x_2$.
        Using the same logic as earlier, for $y \in L_2$ distinct from $a$, the line $yp$ intersects $L_1$ at a point $x$ such that $\phi(x)=y$.
        Finally, since $\phi$ also maps $a \in L_1$ to $a \in L_2$, $\phi$ is a bijection from $L_1$ to $L_2$.

        \underline{Case 2}: $L_1$ does not intersect $L_2$.
        Let $x \in L_1, y \in L_2$, then the line $L_3 \coloneqq xy$ intersects with both $L_1$ and $L_2$, so from case 1 we have bijections
        $\phi_1: L_1 \rightarrow L_3$ and $\phi_2: L_3 \rightarrow L_2$, and thus a bijection $\phi_1 \circ \phi_2: L_1 \rightarrow L_3$.
    \end{proof}

@@

    The Veblen-Young theorem shows finite projective spaces of dimension $n \geq 3$ can always be described using such a field.
    However, finite projective planes are much harder to classify, since they can be constructed in more ways.
    Specifically, planes that do not satisfy the theorem of Desargues.\cite[\S3]{beutelspacher_projective_1998}

    The difficulty in classifying finite projective planes gives rise to an open conjecture:
    \textit{The order of a finite projective plane is always a prime power.}
    The number of non-isomorphic finite projective planes is only known up to order 10: $1, 1, 1, 1, 0, 1, 1, 4, 0$
    due to the computational complexity of verifying this using brute force\cite{lam_computer_1991,lam_non-existence_1989}.
    The closest we have come to a proof of this conjecture is the Bruck-Ryser theorem\cite[p.~26]{beutelspacher_projective_1998} which states if for positive integer $n$,
    $n \equiv 1,2\ (mod\ 4)$ and $n$ is not the sum of two squares, then no plane exists of order $n$.
